\textit{Supporting Information for}: Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories

### 2020-09-04 paper/main.tex

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

\begin{figure}[tp]
  \centering
  \vspace*{-1cm}
  \includegraphics[width=.7\textwidth]{figs/precision_distinctiveness_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  \textbf{B.} Recall distinctiveness by episode event. analogous to Panel A.  \textbf{C.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{D.} Sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 21.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{E.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 21.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

@@

Figure \ref{fig:precision-detail} also similarly illustrates the differences between high and low distinctiveness scores.  We extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) corresponding to the example episode event shown in Panel C (event 21).  We also extracted the annotations for all episode events whose content these participants' single recall events described .  We assigned each episode event a unique color (Fig.~\ref{fig:precision-detail}E), and colored each recall sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 21, whereas the least distinctive recall incorporates content from a much wider range of episode events.

@@

\subsection*{Statistics}
All statistical tests performed in the behavioral analyses were two-sided.  All statistical tests performed in the neural data analyses were two-sided, except for the permutation-based thresholding, which was one-sided.  In this case, we were specifically interested in identifying voxels whose activation time series reflected the temporal structure of the episode and recall trajectories to a \textit{greater} extent than that of the phase-shifted trajectories.
% ------ NOTE: ------
% from the Nature Human Behavior submission requirements: "Specify whether tests were one- or two-tailed, and justify the use of one-tailed tests."
% - Paxton

\subsection*{Modeling the dynamic content of the episode and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the episode comprised 998 hand-generated annotations of short (mean: 2.96s) time segments spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two annotations referring to a break between the first and second scan sessions, during which no fMRI data were collected).  We concatenated the text for all of the annotated features within each segment, creating a ``bag of words'' describing its content and performed some minor preprocessing (e.g., stemming possessive nouns and removing punctuation).  We then re-organized the text descriptions into overlapping sliding windows spanning (up to) 50 annotations each.  In other words, we estimated the ``context'' for each annotated segment using the text descriptions of the preceding 25 annotations, the present annotations, and the following 24 annotations.  To model the context for annotations near the beginning of the episode (i.e., within 25 of the beginning or end), we created overlapping sliding windows that grew in size from one annotation to the full length.  We also tapered the sliding window lengths at the end of the episode, whereby time segments within fewer than 24 annotations of the end of the episode were assigned sliding windows that extended to the end of the episode.  This procedure ensured that each annotation's content was represented in the text corpus an equal number of times.
% ------ NOTE: ------
% need to refine explanation here a little bit
% - Paxton

>>>>

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway;][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

\begin{figure}[tp]
  \centering
  \vspace*{-1cm}
  \includegraphics[width=.7\textwidth]{figs/precision_distinctiveness_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  \textbf{B.} Recall distinctiveness by episode event, analogous to Panel A.  \textbf{C.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{D.} Sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 21.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{E.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 21.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

@@

Figure \ref{fig:precision-detail} also illustrates the differences between high and low distinctiveness scores.  We extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) corresponding to the example episode event shown in Panel C (event 21).  We also extracted the annotations for all episode events whose content these participants' single recall events described .  We assigned each episode event a unique color (Fig.~\ref{fig:precision-detail}E), and colored each recall sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 21, whereas the least distinctive recall incorporates content from a much wider range of episode events.

@@

\subsection*{Statistics}
All statistical tests performed in the behavioral analyses were two-sided.  All statistical tests performed in the neural data analyses were two-sided, except for the permutation-based thresholding, which was one-sided.  In this case, we were specifically interested in identifying voxels whose activation time series reflected the temporal structure of the episode and recall trajectories to a \textit{greater} extent than that of the phase-shifted trajectories.

\subsection*{Modeling the dynamic content of the episode and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the episode comprised 998 hand-generated annotations of short (mean: 2.96s) time segments spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two annotations referring to a break between the first and second scan sessions, during which no fMRI data were collected).  We concatenated the text for all of the annotated features within each segment, creating a ``bag of words'' describing its content and performed some minor preprocessing (e.g., stemming possessive nouns and removing punctuation).  We then re-organized the text descriptions into overlapping sliding windows spanning (up to) 50 annotations each.  In other words, we estimated the ``context'' for each annotated segment using the text descriptions of the preceding 25 annotations, the present annotations, and the following 24 annotations.  To model the context for annotations near the beginning of the episode (i.e., within 25 of the beginning or end), we created overlapping sliding windows that grew in size from one annotation to the full length.  We also tapered the sliding window lengths at the end of the episode, whereby time segments within fewer than 24 annotations of the end of the episode were assigned sliding windows that extended to the end of the episode.  This procedure ensured that each annotation's content was represented in the text corpus an equal number of times.

### 2020-09-04 paper/supporting_information.tex

\title{\textit{Supporting Information for}: Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories}
\author{Andrew C. Heusser\textsuperscript{1, 2, \textdagger}, Paxton C. Fitzpatrick\textsuperscript{1, \textdagger}, and Jeremy R. Manning\textsuperscript{1, *}\\\textsuperscript{1}Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\\textsuperscript{2}Akili Interactive\\Boston, MA 02110\\\textsuperscript{\textdagger}Denotes equal contribution\\\textsuperscript{*}Corresponding author: Jeremy.R.Manning@Dartmouth.edu}

>>>>

\title{\textit{Supporting Information for}: Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories}
\author{Andrew C. Heusser\textsuperscript{1, 2, \textdagger}, Paxton C. Fitzpatrick\textsuperscript{1, \textdagger}, and Jeremy R. Manning\textsuperscript{1, *}\\\textsuperscript{1}Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\\textsuperscript{2}Akili Interactive Labs\\Boston, MA 02110\\\textsuperscript{\textdagger}Denotes equal contribution\\\textsuperscript{*}Corresponding author: Jeremy.R.Manning@Dartmouth.edu}
